Marker imputation efficiency for Genotyping-By-Sequencing data in rice (\emph{Oryza sativa}) and alfalfa (\emph{Medicago sativa}) \thanks{Nelson Nazzicari and Filippo Biscarini contributed equally to the work \\ Corresponding author: Nelson Nazzicari \email{nelson.nazzicari@crea.gov.it}}

### 2016-05-04 Nazzicari_Imputation.tex
aggregato

\subsection{Genotypes}
\label{sec:genotypes}  
GBS detected 32\,706, 40\,734 and 166\,418 SNP loci in Alfalfa-PV, Alfalfa-Med and Rice, respectively (Table~\ref{tab:descriptive_statistics}). The overall missing rate was $66.6\%$, $59.6\%$ and $53.4\%$ in the three datasets. Supplementary Figure S1 shows the density distribution of missing rate per marker in the complete datasets. The amount of missing marker genotypes varied with the threshold of maximum per-marker missing-rate allowed in the data (10\%, 20\%, 40\% and 70\%) and the proportion of artificial missing genotypes that were introduced (1\%, 5\%, 10\% and 20\%). Missing data point counts ranged from a minimum of 3\,878 in Alfalfa-PV with 10\% allowed and 1\% artificial missing genotypes, to a maximum of 5\,580\,616 in Rice with 70\% allowed and 20\% artificial missing genotypes. Over all datasets and allowed/artificial missing genotypes thresholds, the amount of missing data points to be imputed averaged 249\,405.\\
Minor Allele Frequency (MAF) was 0.172, 0.170 and 0.140 in Alfalfa-PV, Alfalfa-Med and Rice, respectively. The three datasets were therefore unbalanced with respect to genotype classes: Rice had the lowest MAF, while Alfalfa-PV and Alfalfa-Med had the smallest minor classes (minor homozygotes 3.5\% and 3.2\% of the total genotypes). Figure~\ref{fig:genotype_classes} reports the proportion of genotypes in each class in the complete datasets (100\%) and as a function of the maximum allowed missing rate per marker. The relative proportion of genotype classes remained approximately stable in rice. In contrast, the heterozygous class in alfalfa tended to become smaller with increasing proportion of allowed genotypes. This reflected the different GBS calling criteria for homozygous and heterozygous SNP: heterozygous SNPs required fewer overlapping reads to be called, while the implemented quality filters on alfalfa required a larger number of reads to call a homozygous locus. Therefore homozygous loci tended to have higher missing rates, and to
be included progressively more frequently with more relaxed thresholds on allowed missing level.
The average missing-rate and MAF in the 12 rice chromosomes were comparable to the whole-rice dataset.\\
Alignment on reference genomes assigned a position to 88.66\%, 57.54\% and 57.86\% of Rice, Alfalfa-Med and Alfalfa-PV markers, respectively, reflecting the difference between having a reference genome available for the exact same species (rice) or only for a close one (alfalfa).

dsahe different proportion of unaligned markers directly reflect the .

@@

Beagle outperformed all other imputation methods in rice, making efficient use of marker position (Figure~\ref{fig:accuracy_rice}). When markers were randomly reshuffled (thus offsetting position information), though, general imputation methods (other than MNI) showed a higher imputation accuracy than Beagle. The accuracy loss with shuffled markers is exacerbated in alfalfa datasets, and in particular in heterozygous and minor homozygous genotypes.\\
Similarly, FILLIN resulted in a 5-8\% accuracy drop when marker positions were shuffled. FILLIN performances were in general lower than other methods due to the fraction of unimputed genotypes, here accounted as imputation errors. On average, FILLIN did not impute 14.6\% of missing data in Rice and practically the entirety of Alfalfa data. Even trying to relax FILLIN parameters, it was impossible to obtain sensible results in alfalfa. Thus, FILLIN accuracy results are not reported in Figure~\ref{fig:accuracy_alfalfa}.

@@

\subsection{Minor Allele Frequency and data (un)balancedness}
\label{sec:maf}
GBS data pose a greater challenge than SNP-array data to imputation algorithms, mainly as a consequence of the much larger quantity of missing genotypes they contain. On our data sets we found missing rates varied from 53\% to 67\%. With larger amounts of missing data the complexity of the imputation problem increases. Imputation errors can have a negative impact on successive analyses (e.g. on genomic predictions \cite{rutkoski_imputation_2013,Annicchiarico2015}).\\
The imputation of missing SNP genotypes is a special case of the broader family of classification problems: three-class missing genotypes (AA, AB and BB) at any given SNP locus are classified based on known genotypes at all remaining data points. Classification problems are known to be harder when data are unbalanced, i.e. the classes appear at different frequencies in the datasets, with typically one over-represented class
(see~\cite{kotsiantis_handling_2006}
and~\cite{sun_classification_2009} for a review). In SNP genotype imputation, data balancedness is directly related to the minor allele frequency (MAF). In the classification of unbalanced observations, it is important to look not only at the total classification accuracy, but also at the per-class accuracy. The total classification accuracy may be misleading, being ``dominated'' by the majority class
\cite{he_learning_2009}.
Indeed, we found that for most methods, even when the total classification accuracy was very high, relatively large error rates were present on minority classes. 
The dependency of imputation results on MAF has already been acknowledged  (e.g.,~\cite{hickey_factors_2012} in maize; \cite{ma_comparison_2013} in cattle; \cite{pei_analyses_2008} in humans). Our per-class dissection of results allowed a deeper insight into the imputation process. Indeed, all imputation algorithms performed considerably better in the majority class rather than in the less frequent classes.
In alfalfa, KNNI was easily the best imputation method in the heterozygous and minor homozygous classes, while SVDI and RFI performed only slightly better than MNI. In rice, Beagle gave the best imputation results, with accuracy close to 100\% both in the major and minor homozygous classes.

\subsection{Missing rate and the ``curse of dimensionality''}
\label{sec:missing_rate_curse_dimensionality}
In general, imputation methods were relatively robust to increasing missing rates. Only KNNI, and to a lesser extent RFI, showed slowly degrading performances at very high missing rates. The divergent response of the different imputation methods to missing rate became apparent only in the most challenging scenarios, when markers with up to 40-70\% missing genotypes were allowed in the data set.\\
KNNI's decrease in imputation accuracy with increasing missing rates can be interpreted in relation to the phenomenon known as the ``curse of dimensionality'' \cite{bellman_dynamic_1957,marimont_nearest_1979}: with the same amount of data, the increasing number of parameters - the ``dimensions'' of the problem - increases and complicates the identification of k neighbors which are close enough to the data point to be classified/imputed. 
On average, the side $l$ of the hypercube to include $k$ neighbors is a function of $k$, $n$ (sample size) and $p$ (number of parameters): $l = \left ( \frac{k}{n} \right )^{1/p}$. With $n$ constant, the hypercube in which the $k$ neighbors lie gets bigger as $p$ increases. This holds especially for local learning methods (e.g.,~K-Nearest Neighbors methods, local regression) which rely heavily on the the information content of the neighborhood. RFI - which only partially relies on local structure of the data - suffered marginally from high missingness. SVD and MNI, which build on more general properties of the data, and Beagle - whose learning process is fundamentally different and specific for genotype imputation - were scarcely affected by missing rates.

\subsection{Imputation efficiency: differences in rice vs. alfalfa}
\label{sec:role_of_species}
Imputation results were significantly different in rice and alfalfa: all imputation algorithms performed consistently better in rice, where Beagle, RFI and KNNI achieved performances comparable to what is reported in literature for SNP-chip data (e.g.,~\cite{vanraden_genomic_2013} for bovine data;
\cite{the_1000_genomes_project_consortium_integrated_2012}
for human data). On the other hand, imputation accuracy in alfalfa data sets was much lower. This can be ascribed to four main factors: \\
i) the imputation problem was simpler in rice than in alfalfa, since the rice dataset comprised only two genotype classes (AA and BB) instead of alfalfa's three (AA, AB, BB); \\
ii) rice is natively diploid while alfalfa, autotetraploid, has been rendered diploid ``in silico'' during SNP calling. This semplification step made alfalfa data less adherent to the underlying biology;\\
iii) rice data have higher marker density (2.4~Kbp/SNP, compared to 24.5~Kbp/SNP for Alfalfa-PV and 19.6~Kbp/SNP for Alfalfa-Med) due to both a higher number of markers and a smaller genome (400~Mb for rice, 800~Mb for Alfalfa). This allowed for higher average values of linkage disequilibrium (LD) among SNP markers, thus facilitating the imputation process;\\
iv) rice markers were aligned on their native genome, while alfalfa markers were aligned on a different species.

>>>>

\subsection{Genotypes}
\label{sec:genotypes}  
GBS detected 32\,706, 40\,734 and 166\,418 SNP loci in Alfalfa-PV, Alfalfa-Med and Rice, respectively (Table~\ref{tab:descriptive_statistics}). The overall missing rate was $66.6\%$, $59.6\%$ and $53.4\%$ in the three datasets. Supplementary Figure S1 shows the density distribution of missing rate per marker in the complete datasets. The amount of missing marker genotypes varied with the threshold of maximum per-marker missing-rate allowed in the data (10\%, 20\%, 40\% and 70\%) and the proportion of artificial missing genotypes that were introduced (1\%, 5\%, 10\% and 20\%). Missing data point counts ranged from a minimum of 3\,878 in Alfalfa-PV with 10\% allowed and 1\% artificial missing genotypes, to a maximum of 5\,580\,616 in Rice with 70\% allowed and 20\% artificial missing genotypes. Over all datasets and allowed/artificial missing genotypes thresholds, the amount of missing data points to be imputed averaged 249\,405.\\
Minor Allele Frequency (MAF) was 0.172, 0.170 and 0.140 in Alfalfa-PV, Alfalfa-Med and Rice, respectively. The three datasets were therefore unbalanced with respect to genotype classes: Rice had the lowest MAF, while Alfalfa-PV and Alfalfa-Med had the smallest minor classes (minor homozygotes 3.5\% and 3.2\% of the total genotypes). Figure~\ref{fig:genotype_classes} reports the proportion of genotypes in each class in the complete datasets (100\%) and as a function of the maximum allowed missing rate per marker. The relative proportion of genotype classes remained approximately stable in rice. In contrast, the heterozygous class in alfalfa tended to become smaller with increasing proportion of allowed genotypes. This reflected the different GBS calling criteria for homozygous and heterozygous SNP: heterozygous SNPs required fewer overlapping reads to be called, while the implemented quality filters on alfalfa required a larger number of reads to call a homozygous locus. Therefore homozygous loci tended to have higher missing rates, and to
be included progressively more frequently with more relaxed thresholds on allowed missing level.
The average missing-rate and MAF in the 12 rice chromosomes were comparable to the whole-rice dataset.\\
Alignment on reference genomes assigned a position to 88.66\%, 57.54\% and 57.86\% of Rice, Alfalfa-Med and Alfalfa-PV markers, respectively, reflecting the difference between having a reference genome available (rice) or using that of a closely related species (alfalfa).

@@

Beagle outperformed all other imputation methods in rice, making efficient use of marker position (Figure~\ref{fig:accuracy_rice}). When markers were randomly reshuffled (thus offsetting position information), though, general imputation methods (other than MNI) showed a higher imputation accuracy than Beagle. The accuracy loss with shuffled markers is exacerbated in alfalfa datasets, and in particular in heterozygous and minor homozygous genotypes.\\
Similarly, FILLIN resulted in a 5-8\% accuracy drop when marker positions were shuffled. FILLIN performances were in general lower than other methods due to the fraction of unimputed genotypes, here accounted as imputation errors. On average, FILLIN did not impute 14.6\% of missing data in Rice and practically the entirety of Alfalfa data. Even trying to relax FILLIN parameters, we couldn't obtain meaningful results in alfalfa. Thus, FILLIN accuracy results are not reported in Figure~\ref{fig:accuracy_alfalfa}.

@@

\subsection{Minor Allele Frequency and data (un)balancedness}
\label{sec:maf}
GBS data pose a greater challenge than SNP-array data to imputation algorithms, mainly as a consequence of the much larger quantity of missing genotypes they contain. In our data sets we found that missing rates varied from 53\% to 67\%. With larger amounts of missing data the complexity of the imputation problem increases. Imputation errors can have a negative impact on successive analyses (e.g. genomic predictions \cite{rutkoski_imputation_2013,Annicchiarico2015}).

The imputation of missing SNP genotypes is a special case of the broader family of classification problems: three-class missing genotypes (AA, AB and BB) at any given SNP locus are classified based on known genotypes at all remaining data points. Classification problems are known to be harder when data are unbalanced, i.e. the classes appear at different frequencies in the datasets, with typically one over-represented class
(see~\cite{kotsiantis_handling_2006}
and~\cite{sun_classification_2009} for a review). In SNP genotype imputation, data balancedness is directly related to the minor allele frequency (MAF). In the classification of unbalanced observations, it is important to look not only at the total classification accuracy, but also at the per-class accuracy. The total classification accuracy may be misleading, being ``dominated'' by the majority class
\cite{he_learning_2009}.
Indeed, we found that for most methods, even when the total classification accuracy was very high, relatively large error rates were present in the minority classes. 
The dependency of imputation results on MAF has already been acknowledged  (e.g.,~\cite{hickey_factors_2012} in maize; \cite{ma_comparison_2013} in cattle; \cite{pei_analyses_2008} in humans). Our per-class dissection of results allowed a deeper insight into the imputation process. Indeed, all imputation algorithms performed considerably better in the majority class rather than in the less frequent classes.
In alfalfa, KNNI was easily the best imputation method in the heterozygous and minor homozygous classes, while SVDI and RFI performed only slightly better than MNI. In rice, Beagle gave the best imputation results, with accuracy close to 100\% both in the major and minor homozygous classes.

\subsection{Missing rate and the ``curse of dimensionality''}
\label{sec:missing_rate_curse_dimensionality}
In general, imputation methods were relatively robust to increasing missing rates. Only KNNI, and to a lesser extent RFI, showed slowly degrading performances at very high missing rates. The divergent response of the different imputation methods to missing rate became apparent only in the most challenging scenarios, when markers with up to 40-70\% missing genotypes were allowed in the dataset.\\
KNNI's decrease in imputation accuracy with increasing missing rates can be interpreted in relation to the phenomenon known as the ``curse of dimensionality'' \cite{bellman_dynamic_1957,marimont_nearest_1979}: with the same amount of data, the increasing number of parameters - the ``dimensions'' of the problem - increases and complicates the identification of k neighbors which are close enough to the data point to be classified/imputed. 
On average, the side $l$ of the hypercube to include $k$ neighbors is a function of $k$, $n$ (sample size) and $p$ (number of parameters): $l = \left ( \frac{k}{n} \right )^{1/p}$. With $n$ constant, the hypercube in which the $k$ neighbors lie gets bigger as $p$ increases. This holds especially for local learning methods (e.g.,~K-Nearest Neighbors methods, local regression) which rely heavily on the the information content of the neighborhood. RFI - which only partially relies on local structure of the data - suffered marginally from high missingness. SVD and MNI, which build on more general properties of the data, and Beagle - whose learning process is fundamentally different and specific for genotype imputation - were scarcely affected by missing rates.

\subsection{Imputation efficiency: differences in rice vs. alfalfa}
\label{sec:role_of_species}
Imputation results were significantly different in rice and alfalfa: all imputation algorithms performed consistently better in rice, where Beagle, RFI and KNNI achieved performances comparable to what is reported in literature for SNP-chip data (e.g.,~\cite{vanraden_genomic_2013} for bovine data;
\cite{the_1000_genomes_project_consortium_integrated_2012}
for human data). On the other hand, imputation accuracy in alfalfa was much lower. This can be ascribed to four main factors: \\
i) the imputation problem was simpler in rice than in alfalfa, since the rice dataset comprised only two genotype classes (AA and BB) instead of alfalfa's three (AA, AB, BB); \\
ii) rice is natively diploid while alfalfa, autotetraploid, has been rendered diploid ``in silico'' during SNP calling. This simplification step made alfalfa data less adherent to the underlying biology;\\
iii) rice data have higher marker density (2.4~Kbp/SNP, compared to 24.5~Kbp/SNP for Alfalfa-PV and 19.6~Kbp/SNP for Alfalfa-Med) due to both a higher number of markers and a smaller genome (400~Mb for rice, 800~Mb for Alfalfa). This allowed for higher average values of linkage disequilibrium (LD) among SNP markers, thus facilitating the imputation process;\\
iv) rice markers were aligned on their native genome, while alfalfa markers were aligned on the genome of a different species.
